Statistical framework for whole-genome bisulfite-sequencing data

### 2014-12-10 Rmarkdown/wgbs_statistical_framework.Rmd
Editing

Beginning with a single sample, I explain the various sources of variation in this statistical model and introduce the mathematical notation that I use throughout my thesis. I then extend this framework to multiple samples. Finally, I describe key variables, common estimators of these and their statistical properties by analysing 51 whole-genome bisulfite-sequencing samples.

@@

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/bs-seq_protocol.pdf}
\caption{A schematic illustrating several sources of variation in a bisulfite-sequencing experiment. Each sample starts as a population of cells, with potentially different methylation patterns. Each DNA fragment is coloured by its originating cell, although this is unknown in practice. Illustrated are PCR amplication bias (unequal representation of DNA from each cell following PCR amplification), sampling variation (not all DNA fragments are sequenced), alignment errors (not all sequenced fragments are mapped or may be mapped to the wrong location, as is the case for the green fragment), and filtering during methylation calling (not reads are used, the blue read, and reads may have read-positions removed, the black reads).}
\label{fig:bs-seq_protocol}
\end{figure}

@@

A methylation locus is a single cytosine, that is, a CpG, CHG or CHH. The set of these loci is labelled $\mathcal{I} = \{pos_{i}: i = 1, \ldots, N_{loci} \}$, where $pos_{i}$ is the genomic co-ordinates of the $i^{th}$ locus, e.g., chr1:+:666. Generally, the number of methylation loci, $N_{loci}$, is not known, although estimates may be made based from the sample or a reference genome. Not knowing the exact value of $N_{loci}$ is no great concern.

@@

$\mathcal{H}_{(i, i + 1, i + m - 1)}$ denotes the set of all fragments containing the m-tuple $(i, i + 1, i + m - 1)$ and $\mathcal{R}_{(i, i + 1, i + m - 1)}$ denotes the set of all reads containing the m-tuple $(i, i + 1, i + m - 1)$.

@@

1. There may be gaps due to the insert size being longer than the sum of the read lengths, e.g. $(i, i + 1, i + 3, i + 4))$. In effect, we have missing data for any intervening methylation loci; the $(i + 2)^{th}$ loci in this example.
2. Loci may be measured twice if the insert size is less than the sum of the read lengths, e.g. read_1 gives us $(i, i + 1)$ and read_2 gives us $(i + 1, i + 2, i + 3)$. In this example we must use only one of read_1 or read_2 as the measurement of the $(i + 1)^{th}$ locus because otherwise we are 'double-counting'.

@@

\begin{equation*}
z: z \in \mathcal{R}_{(i, i + 1, \ldots, i + \text{m} - 1)} = \left\{
  \begin{array}{l l}
    (0, 0, \ldots, 0) & \quad \text{if unmethylated at the } \\
    & \quad \Big ( i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 1)^{th} \Big) \text{ locus}\\
    (0, 0, \ldots, 1) & \quad \text{if unmethylated at the } \\
    \quad \Big (i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 2)^{th} \Big) \text{ locus} \\
      & \quad \text{and methlyated at the } (i + \text{m} - 1)^{th} \text{ locus} \\
    \vdots \\
    (1, 1, \ldots, 1) & \quad \text{if methylated at the } \\
    & \quad \Big (i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 1)^{th} \Big ) \text{ locus}
      \end{array} \right.
\end{equation*}

Note that we do not know from which haplotype ($h$) each read came from, only that all methylation loci in the read came from the same DNA fragment.

@@

A fourth level is how the sample's relate to one another, such as the treatment group of each sample. This fourth level might be defined up-front, such as in a designed experiment looking for differences in methylation between tumour and normal tissue. Conversely, the aim of the experiment might be to _discover_ this grouping, such as in a clustering analysis.

@@

In addition to the complications of the Section \ref{sec:one_sample_complications}, we now have sample-to-sample variation that must be addressed by this framework.

@@

The simplest estimator of $B_{i}$ is $\beta_{i} = \frac{m_{i}}{m_{i} + u_{i}}$, which has been widely used \citep[e.g.,]{Cokus:2008fc, Lister:2008bh, Lister:2009hy, Lister:2011kg}). The values of $m_{i}$ and $u_{i}$ are obtained by methylation calling (see Section \ref{sec:methylation_calling}).

@@

Recently, more sophisticated methods have been proposed to estimate or model the average methylation level. These methods, which are based on $m_{i, j}$ and $u_{i, j}$, include beta-binomial models \citep{Feng:2014iq, Sun:2014fk, Dolzhenko:2014bo}, and smoothing-methods \citep{Hansen:2011gu, Hansen:2012gr, Hebestreit:2013ko}.

@@

`BSmooth`, published in \citet{Hansen:2011gu, Hansen:2012gr} and available in the R/Bioconductor package `bsseq`, and `BiSeq`, published in \citet{Hebestreit:2013ko} and available in the R/Bioconductor package `BiSeq`, take a different approach to getting improved estimates of the $B_{i, j}$. Both `bsseq` and `BiSeq` use statistical smoothing of the 'raw' $\beta_{i, j} = \frac{m_{i, j}}{m_{i, j} + u_{i, j}}$. Smoothing is motivated and justified by the fact that the $B_{i, j}$ are spatially correlated within a sample, i.e., because of the presence of co-methylation (discussed in Chapters \ref{chap:co-methylation} and \ref{chap:co-methylation_review}).

@@

Both `bsseq` and `BiSeq` use a binomial local likelihood smoother. In each case this smoother was chosen because `BSmooth` and `BiSeq` model the number of methylated reads at the $i^{th}$ locus in the $j^{th}$ sample by ($M_{i, j} | n_{i, j} \eqd Binom(n_{i, j}, B_{i, j})$. The smoothing is 'local' to exploit co-methylation, which is a local phenomenon.

@@

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaDensity.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data. Densities were normalised so that the maximum value for each sample is 1. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included. 'Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:EPISCOPE_plotBetaDensity}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/Lister_plotBetaDensity.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Lister data. Densities were normalised so that the maximum value for each sample is 1. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included. 'Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:Lister_plotBetaDensity}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaDensity.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Seisenberger data. Densities were normalised so that the maximum value for each sample is 1. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included. 'Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:Seisenberger_plotBetaDensity}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaDensity.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Ziller data. Densities were normalised so that the maximum value for each sample is 1. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included. 'Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:Ziller_merged_plotBetaDensity}
\end{figure}

@@

These plots show that CpG islands have a more strictly bimodal distribution than do the non-islands. While the majority of CpGs in CpG islands are unmethylated, there are a subset of methylated CpGs in CpG islands in each sample, except for the _E16.5\_male\_1_ sample. The _H1\_r1_ and _H1\_r2_ samples, replicates of an embryonic stem cell line, stand out for having CpGs in CpG islands being more methylated than unmethylated. Theseplots also show that most of the intermediate methylation occurs outside of the CpG islands.

@@

CpG $\beta$-values are often computed by aggregating the $m$ and $u$ counts across the forward and reverse strands. On average, this doubles the sequencing coverage of each CpG but presupposes that the two strands are indeed symmetrically methylated. To investigate the validity of this assumption we can compute the correlation of strand-specific $\beta$-values. Tables \ref{tab:EPISCOPE_strand_correlations}, \ref{tab:Lister_strand_correlations}, \ref{tab:Seisenberger_strand_correlations}, \ref{tab:Ziller_merged_strand_correlations} report the Pearson correlation of these strand-specific $\beta$-values for varying sequencing coverage cutoffs.

@@

The embryonic stem cell samples --- _H1\_r1_, _H1\_r2_ and _HSF1_ from the Lister dataset --- all have less correlated strand-specific $\beta$-values, $r = 0.5 - 0.7$. The other embryonic stem cell samples have higher correlations of $\beta$-values across strands, although the estimates are quite different between the two replicates of the same cell line (_H9_: $r = 0.77$, _H9\_Laurent_: $r = 0.92$). This suggests that it may not be a good idea to combine CpG methylation levels across strands for embryonic stem cell samples.

@@

In papers about DNA methylation there is often a sentence like this quote from \citet{Laird:2003kw}: "About 70% of the CpG dinucleotides in the mammalian genome are methylated". In the context of a whole-genome bisulfite-sequencing experiment, this can be interpreted as the expected $\beta$-value of a randomly selected CpG. However, as can be seen from Figures \ref{EPISCOPE_plotBetaFreqPoly, Lister_plotBetaFreqPoly, Seisenberger_plotBetaFreqPoly, Ziller_plotBetaFreqPoly}, the extreme bimodality of the $\beta$-value distributions means that the expected value is not a particularly useful estimate of the methylation level of a particular CpG. To make a useful statement about the methylation level of a particular CpG really requires that we know whether it is within a CpG island.

@@

\citet{Lister:2008bh} introduced a method "to identify the presence of a methylated cytosine" (detailed in the supplementary material). In the language of these papers, a "methylcytosine" is a cytosine in the reference genome where "at least s [sic; I believe this should be "a"] subset of the genomes within the sample were methylated" \citep[supplementary material]{Lister:2009hy}. This amounts to testing the hypothesis $H_{0}: \beta_{i, j} = 0$ against the one-sided alternative $H_{1}: \beta_{i, j} > 0$. This can be thought of as testing the null hypothesis that the observed number of methylated reads at the $i^{th}$ cytosine were simply due to 'error', where the 'error' is a combination of the estimated sequencing error and the estimated bisulfite-converstion error.

Although the exact procedure is not particularly well described in any of \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}[^mathematical_description], nor is any code made available, I believe the method is as follows. For each cytosine compute a P-value, $P_{i} = \sum_{k = m_{i} + 1}^{k = m_{i} + u_{i}} Prob(X = k)$, where $X = Binom(u_{i} + m_{i}, \epsilon))$ and $\epsilon$ is the estimated 'error'. Any site with an FDR-adjusted P-value below a threshold was declared a 'methylcytosine'[^fdr]

@@

The proportion of cytosines that are identified by this procedure as 'methylcytosines' is a poor estimator of the probability that a cytosine randomly sampled from a haploid genome is methylated, unless the sample is incredibly homogeneous. For example, suppose we had a sample where the true methylation level of every CpG, $B_{i}$, was $0.2$. Given sufficient sequencing coverage, every CpG in the genome would be declared a 'methylcytosine' when in fact for any haploid copy of the genome only $20\%$ of CpGs would be expected to be methylcytosines.

Furthermore, referring to CpGs as 'methylcytosines' results in a loss of information since two 'methylcytosines' may have very different $\beta$-values. For example, in \citet[supplementary figure 2a of]{Lister:2009hy} the authors use a Venn diagram to compare the number of 'methylcytosines' called in two biological replicates to summarise the concordance between the two biological replicates. A far better summary of the biological replicability is to plot the $\beta$-values from each replicate against one another as a scatter plot, as this includes the magnitude of the $\beta$-values and not just whether they are statistically different from zero.

@@

1. The estimate of the standard error depends on the estimate of the mean (i.e., $\beta$), through $se(\beta) = \sqrt{\frac{\beta (1 - \beta)}{m + u}}$. Taking the derivative of this, we see that the maximum standard error, $\sqrt(\frac{0.25}{m + u})$, occurs at $\beta = 0.5$ and the minimum standard error, $0$, occurs at $\beta = 0, 1$.
2. We need to more than just the $\beta$-value to have a sense of how precise an estimate it is. Essentially, we need to also know the sequencing coverage of the methylation loci. Consider two CpGs, one with $m = 1, u = 3$ and the other with $m = 100$ $u = 300$. Both CpGs have $\beta = 1/4$ but the second CpG is measured with much greater precision. Assuming the binomial model, the first CpG has $se(\beta) = \sqrt{\frac{1/4 \times 3/4}{4}} = 0.22$ whereas the second CpG has $se(\beta) = \sqrt{\frac{1/4 \times 3/4}{400}} = 0.02$.
3. Proportions are bound between 0 and 1, inclusive.

@@

One transformation that remains popular, at least in the analysis of DNA methylation microarray data, is the logit-transformation, also known as $\mathcal{M}$-values. An $\mathcal{M}$-value is defined as $logit_{2}(\beta) = \log_{2}(\frac{\beta}{1 - \beta} = \log_{2}\Big ( \frac{m + \alpha}{u + \alpha} \Big )$, where here $m$ and $u$ are the intensities from the methylated and unmethylated probes, respectively, and $\alpha$ is an offset to avoid a numerator or denominator that is zero. Similarly defined $\mathcal{M}$-values, also known as log-ratios, are widely and successfully used in analyses of RNA expression two-colour microarrays \citep[e.g., ]{Smyth:2005ta}.

>>>>

Beginning with a single sample, I explain the various sources of variation in this statistical model and introduce the mathematical notation that I use throughout my thesis. I then extend this framework to multiple samples.

Finally, I describe key variables, common estimators of these and their statistical properties by analysing __HOW MANY__ whole-genome bisulfite-sequencing samples.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/bs-seq_protocol.pdf}
\caption{A schematic illustrating several sources of variation in a bisulfite-sequencing experiment. Each sample starts as a population of cells, with potentially different methylation patterns. Each DNA fragment is coloured by its originating cell, although this is unknown in practice. Illustrated are PCR amplication bias (unequal representation of DNA from each cell following PCR amplification), sampling variation (not all DNA fragments are sequenced), alignment errors (not all sequenced fragments are mapped or may be mapped to the wrong location, as is the case for the green fragment), and filtering during methylation calling (not reads are used, the blue read, and reads may have read-positions removed, the black reads).}
\label{fig:bs-seq_protocol}
\end{figure}

@@

A methylation locus is a single cytosine, that is, a CpG, CHG or CHH. The set of these loci is labelled $\mathcal{I} = \{pos_{i}: i = 1, \ldots, N_{loci} \}$, where $pos_{i}$ is the genomic co-ordinates of the $i^{th}$ locus, e.g., `chr1:+:666`. Generally, the number of methylation loci, $N_{loci}$, is not known, although estimates may be made based from the sample or a reference genome. Not knowing the exact value of $N_{loci}$ is no great concern.

@@

$\mathcal{H}_{(i, i + 1, i + m - 1)}$ denotes the set of all fragments containing the m-tuple $(i, i + 1, i + m - 1)$.

@@

1. There may be gaps due to the insert size being longer than the sum of the read lengths, e.g., $(i, i + 1, i + 3, i + 4))$. In effect, we have missing data for any intervening methylation loci, the $(i + 2)^{th}$ loci in this example.
2. Loci may be measured twice if the insert size is less than the sum of the read lengths, e.g. _read 1_ gives us $(i, i + 1)$ and _read 2_ gives us $(i + 1, i + 2, i + 3)$. In this example we must use only one of _read 1_ or _read 2_ as the measurement of the $(i + 1)^{th}$ locus because otherwise we are 'double-counting'.

@@

\begin{equation*}
z: z \in \mathcal{R}_{(i, i + 1, \ldots, i + \text{m} - 1)} = \left\{
  \begin{array}{l l}
    (0, 0, \ldots, 0) & \quad \text{if unmethylated at the } \\
    & \quad \Big ( i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 1)^{th} \Big) \\
    & \quad \text{ locus}\\
    (0, 0, \ldots, 1) & \quad \text{if unmethylated at the } \\
    & \quad \Big (i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 2)^{th} \Big) \\
    & \quad \text{ locus} \\
      & \quad \text{and methlyated at the } (i + \text{m} - 1)^{th} \text{ locus} \\
    \vdots \\
    (1, 1, \ldots, 1) & \quad \text{if methylated at the } \\
    & \quad \Big (i^{th}, (i + 1)^{th}, \ldots, (i + \text{m} - 1)^{th} \Big ) \\
    & \quad \text{ locus}
      \end{array} \right.
\end{equation*}

Note that we do not know from which haplotype, $h$, each read came from, only that all methylation loci in the read came from the same DNA fragment.

@@

A fourth level is how the sample's relate to one another, such as the treatment group of each sample. This fourth level might be defined up-front, such as in a designed experiment looking for differences in methylation between tumour and normal tissue. Alternatively, the aim of the experiment might be to _discover_ this grouping, such as in a clustering analysis.

@@

In addition to the complications of the Section \ref{sec:one_sample_complications}, we now have sample-to-sample variation that must be addressed within this framework.

@@

The simplest estimator of $B_{i}$ is $\beta_{i} = \frac{m_{i}}{m_{i} + u_{i}}$, which has been widely used \citep[e.g.,][]{Cokus:2008fc, Lister:2008bh, Lister:2009hy, Lister:2011kg}). The values of $m_{i}$ and $u_{i}$ are obtained by methylation calling (see Section \ref{sec:methylation_calling}).

@@

Recently, more sophisticated methods have been proposed to estimate or model the average methylation level. These methods, which are still based on $m_{i, j}$ and $u_{i, j}$, include beta-binomial models \citep{Feng:2014iq, Sun:2014fk, Dolzhenko:2014bo}, and smoothing-methods \citep{Hansen:2011gu, Hansen:2012gr, Hebestreit:2013ko}.

@@

`BSmooth`, published in \citet{Hansen:2011gu, Hansen:2012gr} and available in the R/Bioconductor package `bsseq`, and `BiSeq`, published in \citet{Hebestreit:2013ko} and available in the R/Bioconductor package `BiSeq`, take a different approach to getting improved estimates of the $B_{i, j}$. Both `bsseq` and `BiSeq` use statistical smoothing of the 'raw' $\beta_{i, j} = \frac{m_{i, j}}{m_{i, j} + u_{i, j}}$. Smoothing is motivated and justified by the fact that the $B_{i, j}$ are spatially correlated within a sample, i.e., because of the presence of co-methylation (discussed in Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}).

@@

Both `bsseq` and `BiSeq` use a binomial local likelihood smoother. In each case this smoother was chosen because `BSmooth` and `BiSeq` model the number of methylated reads at the $i^{th}$ locus in the $j^{th}$ sample by $M_{i, j} | n_{i, j} \eqd Binom(n_{i, j}, B_{i, j})$. The smoothing is 'local' to exploit co-methylation, which is a local phenomenon.

@@

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaDensity.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data. Densities were normalised so that the maximum value for each sample is 1. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included. `Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:EPISCOPE_plotBetaDensity}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/Lister_plotBetaDensity.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Lister data. Densities were normalised so that the maximum value for each sample is 1. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included. `Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:Lister_plotBetaDensity}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaDensity.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Seisenberger data. Densities were normalised so that the maximum value for each sample is 1. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included. `Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:Seisenberger_plotBetaDensity}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaDensity.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Ziller data. Densities were normalised so that the maximum value for each sample is 1. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included. `Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:Ziller_merged_plotBetaDensity}
\end{figure}

@@

These plots show that CpG islands have a more strictly bimodal distribution than do the non-islands. While the majority of CpGs in CpG islands are unmethylated, there are a subset of methylated CpGs in CpG islands in each sample, except for the _E16.5\_male\_1_ sample. The _H1\_r1_ and _H1\_r2_ samples, replicates of an embryonic stem cell line, stand out for having CpGs in CpG islands being more methylated than unmethylated. These plots also show that most of the intermediate methylation occurs outside of the CpG islands.

@@

CpG $\beta$-values are often computed by aggregating the $m$ and $u$ counts across the forward and reverse strands. On average, this doubles the sequencing coverage of each CpG but presupposes that the two strands are indeed symmetrically methylated. To investigate the validity of this assumption we can compute the correlation of strand-specific $\beta$-values. Tables \ref{tab:EPISCOPE_strand_correlations}, \ref{tab:Lister_strand_correlations}, \ref{tab:Seisenberger_strand_correlations} and \ref{tab:Ziller_merged_strand_correlations} report the Pearson correlation of these strand-specific $\beta$-values for varying sequencing coverage cutoffs.

@@

Some of the embryonic stem cell samples --- _H1\_r1_, _H1\_r2_ and _HSF1_ from the Lister dataset --- all have less correlated strand-specific $\beta$-values, $r = 0.5 - 0.7$. The other embryonic stem cell samples have higher correlations of $\beta$-values across strands, although the estimates are quite different between the two replicates of the same cell line (_H9_: $r = 0.77$, _H9\_Laurent_: $r = 0.92$). This suggests that it may not be a good idea to combine CpG methylation levels across strands for embryonic stem cell samples.

@@

In papers about DNA methylation there is often a sentence like this quote from \citet{Laird:2003kw}: "About 70% of the CpG dinucleotides in the mammalian genome are methylated". In the context of a whole-genome bisulfite-sequencing experiment, this can be interpreted as the expected $\beta$-value of a randomly selected CpG. However, as can be seen from Figures \ref{EPISCOPE_plotBetaFreqPoly}, \ref{Lister_plotBetaFreqPoly}, \ref{Seisenberger_plotBetaFreqPoly} and \ref{Ziller_plotBetaFreqPoly}, the extreme bimodality of the $\beta$-value distributions means that the expected value is not a particularly useful estimate of the methylation level of a particular CpG. To make a useful statement about the methylation level of a particular CpG really requires that we know whether it is within a CpG island.

@@

\citeauthor{Lister:2008bh} introduced a method "to identify the presence of a methylated cytosine" \citep[Suppmenetary Material]{Lister:2008bh}. In the language of these papers, a "methylcytosine" is a cytosine in the reference genome where "at least s [sic; I believe this should be "a"] subset of the genomes within the sample were methylated" \citep[Supplementary Material]{Lister:2009hy}. This amounts to testing the hypothesis $H_{0}: \beta_{i, j} = 0$ against the one-sided alternative $H_{1}: \beta_{i, j} > 0$. This can be thought of as testing the null hypothesis that the observed number of methylated reads at the $i^{th}$ cytosine were simply due to 'error', where the 'error' is a combination of the estimated sequencing error and the estimated bisulfite-converstion error.

Although the exact procedure is not particularly well described in any of \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}[^mathematical_description], nor is any code made available, I believe the method is as follows. For each cytosine they compute the probability of observing more than $m_{i}$ methylated reads by chance, $P_{i} = \sum_{k = m_{i} + 1}^{k = m_{i} + u_{i}} Pr(X = k)$, where $X = Binom(u_{i} + m_{i}, \epsilon))$ and $\epsilon$ is the estimated 'error'. Any site with an FDR-adjusted P-value below a threshold was declared a "methylcytosine"[^fdr]

@@

The proportion of cytosines that are identified by this procedure as "methylcytosines" is a poor estimator of the probability that a cytosine randomly sampled from a haploid genome is methylated, unless the sample is incredibly homogeneous. For example, suppose we had a sample where the true methylation level of every CpG, $B_{i}$, was $0.2$. Given sufficient sequencing coverage, every CpG in the genome would be declared a "methylcytosine" when in fact for any haploid copy of the genome only $20\%$ of CpGs would be expected to be methylcytosines.

Furthermore, referring to CpGs as 'methylcytosines' results in a loss of information since two 'methylcytosines' may have very different $\beta$-values. For example, in \citet[Supplementary Figure 2a of][]{Lister:2009hy} the authors use a Venn diagram to compare the number of "methylcytosines" called in two biological replicates to summarise the concordance between the two biological replicates. A far better summary of the biological replicability is to plot the $\beta$-values from each replicate against one another as a scatter plot, as this includes the magnitude of the $\beta$-values and not just whether they are statistically different from zero.

@@

1. The estimate of the standard error depends on the estimate of the mean (i.e., $\beta$), through $se(\beta) = \sqrt{\frac{\beta (1 - \beta)}{m + u}}$. Taking the derivative of this, we see that the maximum standard error, $\sqrt{\frac{0.25}{m + u}}$, occurs at $\beta = 0.5$ and the minimum standard error, $0$, occurs at $\beta = 0, 1$.
2. We need to more than just the $\beta$-value to have a sense of how precise an estimate it is. Essentially, we need to also know the sequencing coverage of the methylation loci. Consider two CpGs, one with $m = 1, u = 3$ and the other with $m = 100, u = 300$. Both CpGs have $\beta = 1/4$ but the second CpG is measured with much greater precision. Assuming the binomial model, the first CpG has $se(\beta) = \sqrt{\frac{1/4 \times 3/4}{4}} = 0.22$ whereas the second CpG has $se(\beta) = \sqrt{\frac{1/4 \times 3/4}{400}} = 0.02$.
3. Proportions are bound between 0 and 1, inclusive.

@@

One transformation that remains popular, at least in the analysis of DNA methylation microarray data, is the logit-transformation, also known as $\mathcal{M}$-values. An $\mathcal{M}$-value is defined as $logit_{2}(\beta) = \log_{2} \Big ( \frac{\beta}{1 - \beta} \Big ) = \log_{2}\Big ( \frac{m + \alpha}{u + \alpha} \Big )$, where here $m$ and $u$ are the intensities from the methylated and unmethylated probes, respectively, and $\alpha$ is an offset to avoid a numerator or denominator that is zero. Similarly defined $\mathcal{M}$-values, also known as log-ratios, are widely and successfully used in analyses of RNA expression two-colour microarrays \citep[e.g., ]{Smyth:2005ta}.
